---
title: "UKBB Regression Analysis"
author: "Uditi Shah"
date: '2024-08-13'
output: html_document
---

## 5_UKBB_Regression_Analysis_Binary_C\_Models

## Install relevant packages:

```{r}

# install.packages("pbapply")
library(pbapply)

# install.packages("ggplot2")
library(ggplot2)

# install.packages("GGally")
library(GGally)

# install.packages("ggplot2")
library(ggplot2)

install.packages("gridExtra")
library(gridExtra)
```

# Load the Datasets

```{r}

geno_covar <- readRDS("ukb_adni_genotypes_58_snps_v2.RDS")
# View(geno_covar)


# print(geno_covar[geno_covar$FID == 1000015, ])
# 
# # column/variable names
# names(geno_covar)
# 
# print(data[data$ID == 1000015, ])

```

# 1 Binary Classification - Demographics - glmoost Model Predictions

```{r}

source('regression.R')

# Test the function

model_name='D_glm'

preds_ukbb <- predict(fit_models[[model_name]], new_data = UKBB_subset, type = "prob") %>%
    bind_cols(predict(fit_models[[model_name]], new_data = UKBB_subset)) %>%
    bind_cols(UKBB_subset)

merged_geno_df <- preds_ukbb %>% inner_join(geno_covar %>% select(-conv, -AGE, -SEX, -FH, -APOE4), by=c('ID'='FID'))

result_D_glm <- computeSumStats( yp_dat=merged_geno_df %>% pull(`.pred_No AD`), 
                           merged_geno_df %>% select(-`.pred_No AD`), 
                           is_binary=F)

# result_FH <- computeSumStats( yp_dat=merged_geno_df %>% pull(FH), 
#                            merged_geno_df %>% select(-`.pred_No AD`), 
#                            is_binary=F)

result_true_labels <- computeSumStats( yp_dat=merged_geno_df %>% pull(conv), 
                           merged_geno_df %>% select(-`.pred_No AD`), 
                          is_binary=T)

print(result_D_glm)


###################################################################################
# D_glm_plot_v1 <- plot(-log10(result_true_labels$P), -log10(result$P))
# print(D_glm_plot_v1)

# Create a data frame from your results
D_glm_df <- data.frame(
  snp_effect_size = -log10(result_true_labels$P),
  D_glm = -log10(result_D_glm$P)
)

# Calculate the correlation
correlation_value <- cor(D_glm_df$snp_effect_size, D_glm_df$D_glm)

# Print the correlation value
# print(paste("Correlation: ", round(correlation_value, 3)))

# Create the plot with an identity line and correlation annotation
D_glm_plot_v2 <- ggplot(D_glm_df, aes(x = snp_effect_size, y = D_glm)) +
  geom_point(color = 'blue', size = 2) + # Add points
  labs(
   title = "",
    x = "SNP Effect Size (based on Actual Labels)",
    y = "Predicted SNP Effect Size (based on D)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  ) +
  annotate("text", x = max(D_glm_df$snp_effect_size) * 0.7, 
           y = max(D_glm_df$D_glm) * 0.9, 
           label = paste("Correlation: ", round(correlation_value, 3)), 
           color = "black", size = 5)

# Display the plot
print(D_glm_plot_v2)


###################################################################################
# Identify indices where P-values are above the threshold, then plot without them
I <- which(result_true_labels$P < 0.0000001)

# D_glm_plot2_v1 <- plot(-log10(result_true_labels$P[-I]), -log10(result$P[-I]))
# print(D_glm_plot2_v1)

# Create a data frame for plotting with filtered values
D_glm_df2 <- data.frame(
  snp_effect_size_filtered = -log10(result_true_labels$P[-I]),
  D_glm_filtered = -log10(result_D_glm$P[-I])
)

# Calculate the correlation
correlation_value <- cor(D_glm_df2$snp_effect_size_filtered, D_glm_df2$D_glm_filtered)


# Create the plot with appropriate labels and styles
D_glm_plot2_v2 <- ggplot(D_glm_df2, aes(x = snp_effect_size_filtered, y = D_glm_filtered)) +
  geom_point(color = 'red', size = 2) + # Add points
  labs(
   title = "",
    x = "SNP Effect Size (based on Actual Labels)",
    y = "Predicted SNP Effect Size (based on D)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  ) +
  annotate("text", x = max(D_glm_df2$snp_effect_size_filtered) * 0.7, 
           y = max(D_glm_df2$D_glm_filtered) * 0.9, 
           label = paste("Correlation: ", round(correlation_value, 3)), 
           color = "black", size = 5)
# Display the plot
print(D_glm_plot2_v2)
```

# 2 Binary Classification - Family History - Logistic Regression Model

```{r}

source('regression.R')

# Test the function

model_name='Fh_glm'

preds_ukbb <- predict(fit_models[[model_name]], new_data = UKBB_subset, type = "prob") %>%
    bind_cols(predict(fit_models[[model_name]], new_data = UKBB_subset)) %>%
    bind_cols(UKBB_subset)

merged_geno_df <- preds_ukbb %>% inner_join(geno_covar %>% select(-conv, -AGE, -SEX, -FH, -APOE4), by=c('ID'='FID'))

result_Fh_glm <- computeSumStats( yp_dat=merged_geno_df %>% pull(`.pred_No AD`), 
                           merged_geno_df %>% select(-`.pred_No AD`), 
                           is_binary=F)

# result_FH <- computeSumStats( yp_dat=merged_geno_df %>% pull(FH), 
#                            merged_geno_df %>% select(-`.pred_No AD`), 
#                            is_binary=F)

result_true_labels <- computeSumStats( yp_dat=merged_geno_df %>% pull(conv), 
                           merged_geno_df %>% select(-`.pred_No AD`), 
                          is_binary=T)

print(result_Fh_glm)


###################################################################################
# Fh_glm_plot_v1 <- plot(-log10(result_true_labels$P), -log10(result$P))
# print(Fh_glm_plot_v1)

# Create a data frame from your results
Fh_glm_df <- data.frame(
  snp_effect_size = -log10(result_true_labels$P),
  Fh_glm = -log10(result_Fh_glm$P)
)

# Calculate the correlation
correlation_value <- cor(Fh_glm_df$snp_effect_size, Fh_glm_df$Fh_glm)

# Print the correlation value
# print(paste("Correlation: ", round(correlation_value, 3)))

# Create the plot with an identity line and correlation annotation
Fh_glm_plot_v2 <- ggplot(Fh_glm_df, aes(x = snp_effect_size, y = Fh_glm)) +
  geom_point(color = 'blue', size = 2) + # Add points
  labs(
   title = "",
    x = "SNP Effect Size (based on Actual Labels)",
    y = "Predicted SNP Effect Size (based on Fh)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  ) +
  annotate("text", x = max(Fh_glm_df$snp_effect_size) * 0.7, 
           y = max(Fh_glm_df$Fh_glm) * 0.9, 
           label = paste("Correlation: ", round(correlation_value, 3)), 
           color = "black", size = 5)

# Display the plot
print(Fh_glm_plot_v2)


###################################################################################
# Identify indices where P-values are above the threshold, then plot without them
I <- which(result_true_labels$P < 0.0000001)

# Fh_glm_plot2_v1 <- plot(-log10(result_true_labels$P[-I]), -log10(result$P[-I]))
# print(Fh_glm_plot2_v1)

# Create a data frame for plotting with filtered values
Fh_glm_df2 <- data.frame(
  snp_effect_size_filtered = -log10(result_true_labels$P[-I]),
  Fh_glm_filtered = -log10(result_Fh_glm$P[-I])
)

# Calculate the correlation
correlation_value <- cor(Fh_glm_df2$snp_effect_size_filtered, Fh_glm_df2$Fh_glm_filtered)


# Create the plot with appropriate labels and styles
Fh_glm_plot2_v2 <- ggplot(Fh_glm_df2, aes(x = snp_effect_size_filtered, y = Fh_glm_filtered)) +
  geom_point(color = 'red', size = 2) + # Add points
  labs(
   title = "",
    x = "SNP Effect Size (based on Actual Labels)",
    y = "Predicted SNP Effect Size (based on Fh)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  ) +
  annotate("text", x = max(Fh_glm_df2$snp_effect_size_filtered) * 0.7, 
           y = max(Fh_glm_df2$Fh_glm_filtered) * 0.9, 
           label = paste("Correlation: ", round(correlation_value, 3)), 
           color = "black", size = 5)
# Display the plot
print(Fh_glm_plot2_v2)


```

# 3 Binary Classification - Demographics + Family History - glmoost Model

```{r}

source('regression.R')

# Test the function

model_name='DFh_glm'

preds_ukbb <- predict(fit_models[[model_name]], new_data = UKBB_subset, type = "prob") %>%
    bind_cols(predict(fit_models[[model_name]], new_data = UKBB_subset)) %>%
    bind_cols(UKBB_subset)

merged_geno_df <- preds_ukbb %>% inner_join(geno_covar %>% select(-conv, -AGE, -SEX, -FH, -APOE4), by=c('ID'='FID'))

result_DFh_glm <- computeSumStats( yp_dat=merged_geno_df %>% pull(`.pred_No AD`), 
                           merged_geno_df %>% select(-`.pred_No AD`), 
                           is_binary=F)

result_true_labels <- computeSumStats( yp_dat=merged_geno_df %>% pull(conv), 
                           merged_geno_df %>% select(-`.pred_No AD`), 
                          is_binary=T)

print(result_DFh_glm)


###################################################################################
DFh_glm_df <- data.frame(
  snp_effect_size = -log10(result_true_labels$P),
  DFh_glm = -log10(result_DFh_glm$P)
)

# Calculate the correlation
correlation_value <- cor(DFh_glm_df$snp_effect_size, DFh_glm_df$DFh_glm)

# Create the plot with an identity line and correlation annotation
DFh_glm_plot_v2 <- ggplot(DFh_glm_df, aes(x = snp_effect_size, y = DFh_glm)) +
  geom_point(color = 'blue', size = 2) + # Add points
  labs(
    title = "",
    x = "SNP Effect Size (based on Actual Labels)",
    y = "Predicted SNP Effect Size (based on DFh)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  ) +
  annotate("text", x = max(DFh_glm_df$snp_effect_size) * 0.7, 
           y = max(DFh_glm_df$DFh_glm) * 0.9, 
           label = paste("Correlation: ", round(correlation_value, 3)), 
           color = "black", size = 5)

# Display the plot
print(DFh_glm_plot_v2)


###################################################################################
I <- which(result_true_labels$P < 0.0000001)

DFh_glm_df2 <- data.frame(
  snp_effect_size_filtered = -log10(result_true_labels$P[-I]),
  DFh_glm_filtered = -log10(result_DFh_glm$P[-I])
)

# Calculate the correlation
correlation_value <- cor(DFh_glm_df2$snp_effect_size_filtered, DFh_glm_df2$DFh_glm_filtered)

DFh_glm_plot2_v2 <- ggplot(DFh_glm_df2, aes(x = snp_effect_size_filtered, y = DFh_glm_filtered)) +
  geom_point(color = 'red', size = 2) + # Add points
  labs(
   title = "",
    x = "SNP Effect Size (based on Actual Labels)",
    y = "Predicted SNP Effect Size (based on DFh)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  ) +
  annotate("text", x = max(DFh_glm_df2$snp_effect_size_filtered) * 0.7, 
           y = max(DFh_glm_df2$DFh_glm_filtered) * 0.9, 
           label = paste("Correlation: ", round(correlation_value, 3)), 
           color = "black", size = 5)
# Display the plot
print(DFh_glm_plot2_v2)

```

# 4 Binary Classification - DGFh - Logistic Regression Model

```{r}

source('regression.R')

# Test the function

model_name='DGFh_glm'

preds_ukbb <- predict(fit_models[[model_name]], new_data = UKBB_subset, type = "prob") %>%
    bind_cols(predict(fit_models[[model_name]], new_data = UKBB_subset)) %>%
    bind_cols(UKBB_subset)

merged_geno_df <- preds_ukbb %>% inner_join(geno_covar %>% select(-conv, -AGE, -SEX, -FH, -APOE4), by=c('ID'='FID'))

result_DGFh_glm <- computeSumStats( yp_dat=merged_geno_df %>% pull(`.pred_No AD`), 
                           merged_geno_df %>% select(-`.pred_No AD`), 
                           is_binary=F)

result_true_labels <- computeSumStats( yp_dat=merged_geno_df %>% pull(conv), 
                           merged_geno_df %>% select(-`.pred_No AD`), 
                          is_binary=T)

print(result_DGFh_glm)


###################################################################################
# DGFh_glm_df <- data.frame(
#   snp_effect_size = -log10(result_true_labels$P),
#   DGFh_glm = -log10(result_DGFh_glm$P)
# )
# 
# # Calculate the correlation
# correlation_value <- cor(DGFh_glm_df$snp_effect_size, DGFh_glm_df$DGFh_glm)
# 
# # Create the plot with an identity line and correlation annotation
# DGFh_glm_plot_v2 <- ggplot(DGFh_glm_df, aes(x = snp_effect_size, y = DGFh_glm)) +
#   geom_point(color = 'blue', size = 2) + # Add points
#   labs(
#    title = "",
#     x = "SNP Effect Size (based on Actual Labels)",
#     y = "Predicted SNP Effect Size (based on DGFh)"
#   ) +
#   theme_minimal() +
#   theme(
#     plot.title = element_text(hjust = 0.5),
#     axis.text = element_text(size = 12),
#     axis.title = element_text(size = 14)
#   ) +
#   annotate("text", x = max(DGFh_glm_df$snp_effect_size) * 0.7, 
#            y = max(DGFh_glm_df$DGFh_glm) * 0.9, 
#            label = paste("Correlation: ", round(correlation_value, 6)), 
#            color = "black", size = 5)
# 
# # Display the plot
# print(DGFh_glm_plot_v2)
# 
# # I'm getting the Nan for the Cor in this plot because there are some P=0 values  causing inf when u take -log and that's why. so get rid of that and use the plot below
# any(is.na(result_true_labels$P)) || any(is.na(result_DGFh_glm$P))
# # [1] FALSE
# any(is.infinite(-log10(result_true_labels$P))) || any(is.infinite(-log10(result_DGFh_glm$P)))
# # [1] TRUE


###################################################################
# Replace zero or extremely small P-values with a small value (e.g., 1e-10)
result_true_labels$P[result_true_labels$P == 0] <- 1e-10
result_DGFh_glm$P[result_DGFh_glm$P == 0] <- 1e-10

# Apply -log10 transformation and handle infinite values
DGFh_glm_df <- data.frame(
  snp_effect_size = -log10(result_true_labels$P),
  DGFh_glm = -log10(result_DGFh_glm$P)
)

# Replace infinite values with NA (or a large number)
DGFh_glm_df$snp_effect_size[is.infinite(DGFh_glm_df$snp_effect_size)] <- NA
DGFh_glm_df$DGFh_glm[is.infinite(DGFh_glm_df$DGFh_glm)] <- NA

# Remove rows with NA values
DGFh_glm_df_clean <- na.omit(DGFh_glm_df)

# Calculate the correlation again on the cleaned data
correlation_value <- cor(DGFh_glm_df_clean$snp_effect_size, DGFh_glm_df_clean$DGFh_glm)

# Create the plot
DGFh_glm_plot_v2 <- ggplot(DGFh_glm_df_clean, aes(x = snp_effect_size, y = DGFh_glm)) +
  geom_point(color = 'blue', size = 2) + # Add points
  labs(
    title = "",
    x = "SNP Effect Size (based on Actual Labels)",
    y = "Predicted SNP Effect Size (based on DGFh)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  ) +
  annotate("text", x = max(DGFh_glm_df_clean$snp_effect_size) * 0.7, 
           y = max(DGFh_glm_df_clean$DGFh_glm) * 0.9, 
           label = paste("Correlation: ", round(correlation_value, 6)), 
           color = "black", size = 5)

# Display the plot
print(DGFh_glm_plot_v2)



###################################################################################
I <- which(result_true_labels$P < 0.0000001)

DGFh_glm_df2 <- data.frame(
  snp_effect_size_filtered = -log10(result_true_labels$P[-I]),
  DGFh_glm_filtered = -log10(result_DGFh_glm$P[-I])
)

# Calculate the correlation
correlation_value <- cor(DGFh_glm_df2$snp_effect_size_filtered, DGFh_glm_df2$DGFh_glm_filtered)

DGFh_glm_plot2_v2 <- ggplot(DGFh_glm_df2, aes(x = snp_effect_size_filtered, y = DGFh_glm_filtered)) +
  geom_point(color = 'red', size = 2) + # Add points
  labs(
   title = "",
    x = "SNP Effect Size (based on Actual Labels)",
    y = "Predicted SNP Effect Size (based on DGFh)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  ) +
  annotate("text", x = max(DGFh_glm_df2$snp_effect_size_filtered) * 0.7, 
           y = max(DGFh_glm_df2$DGFh_glm_filtered) * 0.9, 
           label = paste("Correlation: ", round(correlation_value, 3)), 
           color = "black", size = 5)

# Display the plot
print(DGFh_glm_plot2_v2)


##############################################################


```

# Plot all the plots side by Side

```{r}

RA_BC_plot1 <- grid.arrange(Fh_glm_plot_v2, D_glm_plot_v2, DFh_glm_plot_v2, DGFh_glm_plot_v2, ncol = 2, nrow = 2)

# Save the updated plot
ggsave("RA_BC_plot1.png", plot = RA_BC_plot1, width = 10, height = 10)
```

```{r}

RA_BC_plot2 <- grid.arrange(Fh_glm_plot2_v2, D_glm_plot2_v2, DFh_glm_plot2_v2, DGFh_glm_plot2_v2, ncol = 2, nrow = 2)

# Save the updated plot
ggsave("RA_BC_plot2.png", plot = RA_BC_plot2, width = 10, height = 10)
```

\#########################
